Semantic Linearity in a Lazy Optimising Compiler

### 2025-01-27 icfp.tex
RM new semantics fixes
tweaked semantics

In this section we develop a lazy big-step operational semantics for $\lambda_\Delta^\pi$ in the
style of Launchbury~\cite{10.1145/158511.158618}. The semantics is
equipped with an environment which assigns values to variables and is
mutated in order to express lazy evaluation. Terms are explicit about
their sharing (in the sense of shared reductions in lazy evaluation),
and so we perform a translation step that makes all sharing explicit
through let-binders.
Below we write $\llet{x{:}_{\Delta}\sigma = e,\dots,y{:}_{\Delta'}=e''
}{e'}$ for the iterated let-binding of variables $x , \dots , y$.       
\[
  \begin{array}{lclr}
    x^* & \triangleq & x\\

    (\Lambda p.~e)^* & \triangleq & \Lambda p.~(e)^*\\
    (\lambda x:_\pi\sigma .e)^* & \triangleq & \lambda x:_\pi\sigma . (e)^*\\
    (e~\pi)^* & \triangleq & (e)^*~\pi\\
    (e~x)^* & \triangleq & (e)^*~x\\
    (e~e')^* & \triangleq & \llet{x{:}_{\Delta}\sigma =
                            (e')^*}{(e)^*~x}     \\ &&\qquad \mbox{with
                            $\Gamma;\Delta' \vdash e :
                            \sigma\rightarrow_\pi \varphi$ and
                                                   $\Gamma;\Delta \vdash e' : \sigma$}\\
    K~e_1~\dots~e_n & \triangleq & \llet{x_1:_{\Delta_1} \sigma_1 =
                                   (e_1)^*,\dots,x_n:_{\Delta_n}
                                   \sigma_n =
                                   (e_n)^*}{K~x_1~\dots~x_n} \\ &&
                                                               \qquad\mbox{with
                                                               $\Gamma;\Delta_1\vdash
                                                               e_1 :
                                                               \sigma_1
                                                               \dots \Gamma;\Delta_n\vdash
                                                               e_n :
                                                               \sigma_n$}\\                                              
(\ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\rho\to e'}\}})^* &
                                                               \triangleq
                                &
                                  \ccase{(e)^*}{z{:}_{\Delta}\sigma~\{\overline{\rho\to (e')^*}\}}\\
    (\llet{x{:}_{\Delta}\sigma = e}{e'}      )^* & \triangleq & \llet{x{:}_{\Delta}\sigma  = (e)^*}{(e')^*}    \\
    (\lletrec{\overline{x{:}_{\Delta}\sigma = e}}{e'}  )^* & \triangleq &\lletrec{\overline{x{:}_{\Delta}\sigma = (e)^*}}{(e')^*}  
      \end{array}
\]

@@

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma : \Lambda p.e \Downarrow \Gamma : \Lambda p.e}
    \qquad
    \infer[]
    {\Gamma : e\Downarrow \Delta : \Lambda p.e' \quad \Delta :
    e'[\pi/p]\Downarrow \Theta : v}
    {\Gamma : e~\pi\Downarrow \Theta : v }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma : \lambda x:_\pi \sigma . e \Downarrow \Gamma : \lambda x:_\pi \sigma . e }
    \qquad
    \infer[]
    {\Gamma : e \Downarrow \Delta : \lambda y:_\pi \sigma . e' \quad
    \Delta : e'[x/y] \Downarrow \Theta : v }
    {\Gamma : e~x \Downarrow \Theta : v}\\[1.5em]
    \infer[]
    {\Gamma : e \Downarrow \Delta : v}
    {(\Gamma , x:_\omega \sigma = e) : x \Downarrow (\Delta,x:_\omega
    \sigma = v) : v}
    \qquad
    \infer[]
    {(\Gamma,x:_\omega \sigma = e) : e' \Downarrow \Theta : v}
    {\Gamma : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta : v
    }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma : K~x_1~\dots~x_n \Downarrow \Gamma : K~x_1~\dots~x_n}
    \qquad
    \infer[]
    {\Gamma : e \Downarrow \Delta : K~x_1~\dots~x_n \quad
    \Delta : e'[x_i/y_i][K~x_1~\dots~x_n/z] \Downarrow \Theta : v }
    {\Gamma :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}}
    \Downarrow \Theta : v}\\[1.5em]
    \infer[]
    {\Gamma : e \Downarrow \Delta : K~x_1~\dots~x_n \quad
    \Delta : e'[K~x_1~\dots~x_n/z] \Downarrow \Theta : v }
    {\Gamma :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\} \Downarrow \Theta : v}
    \qquad
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta
    : v
    }
    {\Gamma : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
    \Downarrow \Theta : v}
    \end{array}
\]

@@

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
    \Lambda p.e) : \forall p . \sigma }
    \\[1.5em]
    \infer[]
    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
    \Lambda p.e') : \forall p . \sigma \quad \Gamma ;\Delta \vdash
    (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' : v) : \sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
    \mid v) : \sigma  }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
    \rightarrow_\pi \tau }\\[1.5em]
    \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_\pi \sigma . e') : \sigma \rightarrow_\pi \tau
    \quad
    \Gamma ; \Delta \vdash (\Theta' \mid e'[x/y]) \Downarrow (\Theta''
    : v) : \tau }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta'
    \mid v) : \tau}\\[1.5em]

@@

    \infer[]
    {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
    \Downarrow (\Theta' \mid v) : \tau}
    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta}\sigma =
    e}{e'})\Downarrow (\Theta' \mid v) : \tau
    }\\[1.5em]

    \infer[]
    {\,}
    {\Gamma ; \Delta \vdash (\Theta \mid K~x_1~\dots~x_n) \Downarrow
    (\Theta \mid K~x_1~\dots~x_n) : T}
    \\[1.5em]

    \infer[]
    {\Gamma;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' :
    K~x_1~\dots~x_n) : T \quad
    \Gamma ;\Delta \vdash (\Theta' \mid e'[x_i/y_i][K~x_1~\dots~x_n/z]
    ) \Downarrow (\Theta'' : v) :\tau }
    {\Gamma ;\Delta \vdash (\Theta \mid
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}})
    \Downarrow (\Theta'' \mid  v) : \tau}\\[1.5em]

    \infer[]
    {\Gamma ;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
    K~x_1~\dots~x_n) : T \quad
    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~x_1~\dots~x_n/z])
    \Downarrow (\Theta'' \mid v) : \tau }
    {\Gamma ; \Delta \vdash (\Theta \mid 
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
    \\[1.5em]
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
    e_i} \mid e' )\Downarrow (\Theta'
    \mid v) : \tau
    }
    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
    \Downarrow (\Theta' \mid v) : \tau}

@@

>>>>

In this section we develop a lazy big-step operational semantics for $\lambda_\Delta^\pi$ in the
style of Launchbury~\cite{10.1145/158511.158618}. The semantics is
equipped with an environment which assigns values to variables and is
mutated in order to express lazy evaluation. Terms are explicit about
their sharing (in the sense of shared reductions in lazy evaluation),
and so we perform a translation step that makes all sharing explicit
through let-binders.
Below we write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
let-binding of variables $\ov{x_i}$.
\[
  \begin{array}{lclr}
    x^* & \triangleq & x\\

    (\Lambda p.~e)^* & \triangleq & \Lambda p.~(e)^*\\
    (\lambda x:_\pi\sigma .e)^* & \triangleq & \lambda x:_\pi\sigma . (e)^*\\
    (e~\pi)^* & \triangleq & (e)^*~\pi\\
    (e~x)^* & \triangleq & (e)^*~x\\
    (e~e')^* & \triangleq & \llet{x{:}_{\Delta}\sigma =
                            (e')^*}{(e)^*~x}     \\ &&\qquad \mbox{with
                            $\Gamma;\Delta' \vdash e :
                            \sigma\rightarrow_\pi \varphi$} \\ &&\qquad \mbox{and
                                       $\Gamma;\Delta \vdash e' : \sigma$}\\
      K~\ov{e_i} & \triangleq & \llet{\ov{x_i:_{\Delta_i} \sigma_i =
                                   (e_i)^*}}{K~\ov{x_i}} \\ &&
                                       \qquad\mbox{with
                                       $\ov{\Gamma;\Delta_i\vdash
                                       e_i :
                                       \sigma_i}$}\\                                              
(\ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\rho\to e'}\}})^* &
                                                               \triangleq
                                &
                                  \ccase{(e)^*}{z{:}_{\Delta}\sigma~\{\overline{\rho\to (e')^*}\}}\\
    (\llet{x{:}_{\Delta}\sigma = e}{e'}      )^* & \triangleq & \llet{x{:}_{\Delta}\sigma  = (e)^*}{(e')^*}    \\
    (\lletrec{\overline{x{:}_{\Delta}\sigma = e}}{e'}  )^* & \triangleq &\lletrec{\overline{x{:}_{\Delta}\sigma = (e)^*}}{(e')^*}  
      \end{array}
\]

@@

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma : \Lambda p.e \Downarrow \Gamma : \Lambda p.e}
    \qquad
    \infer[]
    {\Gamma : e\Downarrow \Delta : \Lambda p.e' \quad \Delta :
    e'[\pi/p]\Downarrow \Theta : v}
    {\Gamma : e~\pi\Downarrow \Theta : v }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma : \lambda x:_\pi \sigma . e \Downarrow \Gamma : \lambda x:_\pi \sigma . e }
    \qquad
    \infer[]
    {\Gamma : e \Downarrow \Delta : \lambda y:_\pi \sigma . e' \quad
    \Delta : e'[x/y] \Downarrow \Theta : v }
    {\Gamma : e~x \Downarrow \Theta : v}\\[1.5em]
    \infer[]
    {\Gamma : e \Downarrow \Delta : v}
    {(\Gamma , x:_\omega \sigma = e) : x \Downarrow (\Delta,x:_\omega
    \sigma = v) : v}
    \qquad
    \infer[]
    {(\Gamma,x:_\omega \sigma = e) : e' \Downarrow \Theta : v}
    {\Gamma : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta : v
    }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma : K~\ov{x_i} \Downarrow \Gamma : K~\ov{x_i}}
    \qquad
    \infer[]
    {\Gamma : e \Downarrow \Delta : K~\ov{x_i} \quad
    \Delta : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta : v }
    {\Gamma :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}}
    \Downarrow \Theta : v}\\[1.5em]
    \infer[]
    {\Gamma : e \Downarrow \Delta : K~\ov{x_i} \quad
    \Delta : e'[K~\ov{x_i}/z] \Downarrow \Theta : v }
    {\Gamma :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\} \Downarrow \Theta : v}
    \qquad
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta
    : v
    }
    {\Gamma : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
    \Downarrow \Theta : v}
    \end{array}
\]

@@

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
    \Lambda p.e) : \forall p . \sigma }
    \\[1.5em]
    \infer[]
    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
    \Lambda p.e') : \forall p . \sigma \quad \Gamma ;\Delta \vdash
    (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
    \mid v) : \sigma  }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
    \rightarrow_\pi \tau }\\[1.5em]
    \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_\pi \sigma . e') : \sigma \rightarrow_\pi \tau
    \quad
    \Gamma ; \Delta \vdash (\Theta' , y:_{\pi} \sigma = x \mid e') \Downarrow (\Theta''
    \mid v) : \tau }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau}\\[1.5em]

@@

    \infer[]
    {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
    \Downarrow (\Theta' \mid v) : \tau}
    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta'}\sigma =
    e}{e'})\Downarrow (\Theta' \mid v) : \tau
    }\\[1.5em]

    \infer[]
    {\,}
    {\Gamma ; \Delta \vdash (\Theta \mid K~\ov{x_i}) \Downarrow
    (\Theta \mid K~\ov{x_i}) : T}
    \\[1.5em]

    \infer[]
    {\Gamma;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' :
    K~\ov{x_i}) : T \quad
    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z]
    ) \Downarrow (\Theta'' \mid v) :\tau }
    {\Gamma ;\Delta \vdash (\Theta \mid
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}})
    \Downarrow (\Theta'' \mid  v) : \tau}\\[1.5em]

    \infer[]
    {\Gamma ;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
    K~\ov{x_i}) : T \quad
    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
    \Downarrow (\Theta'' \mid v) : \tau }
    {\Gamma ; \Delta \vdash (\Theta \mid 
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
    \\[1.5em]
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
    e_i} \mid e' )\Downarrow (\Theta'
    \mid v) : \tau
    }
    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
    \Downarrow (\Theta' \mid v) : \tau}

@@

 \begin{definition}[Well-typed State]
   We write $\vdash \Theta \mid e : \tau$ to denote:

   \[
\vdash \llet{\Theta}{e} : \tau
   \]

 \end{definition}

\begin{theorem}[Type Preservation]
  Let $(\Theta \mid e) \Downarrow (\Theta' \mid v) : \tau$.
  If $\vdash \Theta \mid e : \tau$ then $\vdash \Theta' \mid v : \tau$. 
 \end{theorem}
